First paper - Draft

### 2017-06-22 paper/paper.Rnw
Updated countries for graphs + updated graph - first results writeup

  countries <- c("Australia",
                 "Germany",
                 "Denmark",
                 "Spain",
                 "France",
                 "Italy",
                 "Netherlands",
                 "Sweden",
                 "Finland",
                 "United States",
                 "United Kingdom")
@

@@

results_math <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.1, 0.9))
results_read <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.1, 0.9))
results_math_topmid <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.5, 0.9))
results_read_topmid <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.5, 0.9))
results_math_midbottom <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.1, 0.5))
results_read_midbottom <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.1, 0.5))

# results_math <- read_rds("./data/delete.Rdata")
# results_read <- read_rds("./data/delete_read.Rdata")
# results_math_topmid <- read_rds("./data/delete_math_topmid.Rdata")
# results_read_topmid <- read_rds("./data/delete_read_topmid.Rdata")
# results_math_midbottom <- read_rds("./data/delete_math_midbottom.Rdata")
# results_read_midbottom <- read_rds("./data/delete_read_midbottom.Rdata")
# US is missing for reading

@@

enframer <- function(df) {
  df %>%
    enframe() %>%
    unnest()
}

var_years %>%
  setNames(seq(2000, 2015, 3)) %>%
  enframer() %>%
  ggplot(aes(name, value)) + geom_col()
@

@@

We also calculate the standard error of this difference and use it to create uncertainty intervals in the estimation of time trends. We fit a multilevel rather than a linear model because by pooling the information together, we weight countries appropriately to their sample size. Given that including the SES dummy reduces the sample size considerably, we want to be able to estimate each country-difference as accurately as possible.

@@

% Figure out how to ouput data frames as latex tables.
<<sample_size, cache = TRUE>>=

# Get sample counts for each dummy
sample_size_calc <- function(df, probs, selected = F, cnts = NULL) {

@@

Calculate the difference between 90/10 and then the ratio of this with the annual gain of one year of schooling (\citet{micklewright} did it). They also find that France is a lower dispersion country (but in my results this is changing a lot). They also find that Germany and US are very dispersed countries (quite the opposite)

@@

<<graphing_9010gaps, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=

# 90/10 gaps acros countries
complete_data_topbottom %>%
  filter(country %in% c("United States", "Netherlands", "France",
                        "Germany", "Poland", "Finland")) %>%
  ggplot(aes(as.factor(wave), difference, group = type_test, colour = type_test)) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.1) +
  geom_hline(yintercept = 0, linetype = "longdash") +
  geom_line() +
  geom_point(size = 0.5) +
  coord_cartesian(ylim = c(-0.5, 3)) +
  facet_wrap(~ country)
@

>>>>

  countries <- c("Finland",
                 "France",
                 "New Zealand",
                 "Austria",
                 "Australia",
                 "Sweden",
                 "Czech Republic",
                 "Canada",
                 "Hungary",
                 "Iceland", "Netherlands",
                 "Spain",
                 "Belgium",
                 "Italy", "Norway",
                 "United Kingdom",
                 "Greece",
                 "Denmark",
                 "Israel",
                 "Poland",
                 "United States",
                 "Germany",
                 "Turkey",
                 "Russia")
@

@@

# results_math <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.1, 0.9))
# results_read <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.1, 0.9))
# results_math_topmid <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.5, 0.9))
# results_read_topmid <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.5, 0.9))
# results_math_midbottom <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.1, 0.5))
# results_read_midbottom <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.1, 0.5))

results_math <- read_rds("./data/delete.Rdata")
results_read <- read_rds("./data/delete_read.Rdata")
results_math_topmid <- read_rds("./data/delete_math_topmid.Rdata")
results_read_topmid <- read_rds("./data/delete_read_topmid.Rdata")
results_math_midbottom <- read_rds("./data/delete_math_midbottom.Rdata")
results_read_midbottom <- read_rds("./data/delete_read_midbottom.Rdata")
# US is missing for reading

@@

enframer <- function(df, col_name = "name") {
  df %>%
    enframe(name = col_name) %>%
    unnest()
}

var_years %>%
  setNames(seq(2000, 2015, 3)) %>%
  enframer(name) %>%
  ggplot(aes(name, value)) + geom_col()
@

@@

We also calculate the standard error of this difference and use it to create uncertainty intervals in the estimation of time trends. We fit a multilevel rather than a linear model because by pooling the information together, we weight countries appropriately to their sample size. Given that including the SES dummy reduces the sample size considerably, we want to be able to estimate each country-difference as accurately as possible. For further exploration, we also compute the 90th/50th and 50th/10th SES gaps following the same method outlined above.

@@

<<escsdummy_charachteristics, cache = TRUE, eval = F>>=
.x <- adapted_year_data[[1]]
country <- filter(.x, country == "United States")
probs <- c(0.1, 0.9)

escs_dummy_creator <- function(df, probs) {

    map(df, function(.x) {

      .x
    })
}

escs_data <- escs_dummy_creator(adapted_year_data, c(0.1, 0.9))

# Make recode the education variable into three categories
map_lgl(escs_data, ~ "HISEI" %in% names(.x))

escs_data[[1]] %>%
  select(country, escs_dummy, high_edu_broad, ISEI) %>%
  filter(country %in% countries) %>%
  count(country, escs_dummy, high_edu_broad) %>%
  group_by(country, escs_dummy) %>%
  mutate(total_n = sum(n),
         perc = n / total_n * 100) %>%
  select(country, escs_dummy, high_edu_broad, perc) %>%
  map_if(is_double, round, 2) %>%
  as_tibble() %>%
  filter(!is.na(escs_dummy)) %>%
  spread(high_edu_broad, perc)

escs_data[[1]] %>%
  select(country, escs_dummy, high_edu_broad, ISEI) %>%
  filter(country %in% countries) %>%
  count(country, escs_dummy, isei_group = cut_number(ISEI, 4)) %>%
  group_by(country, isei_group) %>%
  mutate(total_n = sum(n),
         perc = n / total_n * 100) %>%
  select(country, escs_dummy, isei_group, perc) %>%
  map_if(is_double, round, 2) %>%
  as_tibble() %>%
  filter(!is.na(escs_dummy)) %>%
  spread(isei_group, perc) %>%
  left_join(m, by = c("country", "escs_dummy"))

% Figure out how to ouput data frames as latex tables.
<<sample_size, cache = TRUE>>=
# Get sample counts for each dummy
sample_size_calc <- function(df, probs, selected = F, cnts = NULL) {

@@

Calculate the difference between 90/10 and then the ratio of this with the annual gain of one year of schooling (\citet{micklewright} did it).

<<graphing_9010gaps, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
# 90/10 gaps acros countries
diff_increase_fun <- function(df) {

      .x %>%
        map_if(is_double, round, 3) %>%
        as_tibble() %>%
        transmute(type_test,
                  country,
                  diff = round(((!!last_year) - (!!first_year)), 1),
                  sd_year = year_sd,
                  diff_lower = diff - 1 * year_sd,
                  diff_upper = diff + 1 * year_sd,
                  years_available = years_available)
    })
  data_ready
}

diff_data <-
  diff_increase_fun(complete_data_topbottom) %>%
  enframer("country") %>%
  filter(country %in% countries) %>%
  select(country, type_test, diff) %>%
  split(.$type_test) %>%
  map(~ .x %>% select(-type_test) %>% deframe())

lm_data <- function(df) {
  lm(log(difference) ~ wave, data = df) %>%
    broom::tidy() %>%
    mutate(estimate = exp(estimate))
}

ordered_cnt <-
  complete_data_topbottom %>%
  filter(type_test == "math") %>%
  select(wave, country, difference) %>%
  filter(country %in% countries) %>%
  split(.$country) %>%
  map(lm_data) %>%
  enframer("country") %>%
  filter(term == "wave") %>%
  arrange(-estimate) %>%
  pull(country)

complete_data_topbottom %>%
  filter(country %in% countries) %>%
  mutate(country = factor(country, levels = ordered_cnt, ordered = TRUE)) %>%
  ggplot(aes(as.factor(wave), difference, group = type_test, colour = type_test)) +
  geom_line(stat = "smooth", method = "lm", aes(group = 1),
            formula = y ~ splines::ns(x, 3), linetype = "longdash",
            colour = "black") +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.1, alpha = 0.4) +
  geom_hline(yintercept = 0, linetype = "longdash") +
  geom_line(alpha = 0.4) +
  geom_point(size = 0.5, alpha = 0.4) +
  coord_cartesian(ylim = c(0, 3)) +
  facet_wrap(~ country)
@

We start by look at the achievement for some countries. We plot both mathematics and reading gaps for each country and also a cubic trend spline for both test scores pooled. Some countries have increased their achievement very strongly. For example, France, Austria and surprisingly Sweden. In mathematics, France increase the gap by roughly \Sexpr{diff_data$math[["France"]]}. This increase happens similarly for literacy with an increase of \Sexpr{diff_data$read[["France"]]}. For such a short period of time, the magnitud of the increase is reasonably big.

Given that no one has estimated the evolution of the gap we can't cross-check how other empirical estimations put France. However, we can make some educated inferences.  For example, \citet{micklewright} also finds that France is a low dispersion country, specifically in the year 2000. They also find that Germany and US are very dispersed countries (quite the opposite).

@@

### 2017-06-22 paper/paper.tex
Updated countries for graphs + updated graph - first results writeup

In order the study the above mentioned questions I will be using the Programme for International Student Assessment (PISA). PISA is a survey carried out every three years that aims to evaluate education systems by testing the skills and knowledge of 15-year-old students. Currrently, PISA has six waves starting in the year 2000 up until 2015, where recently, over half a million students were tested in mathematics, literacy and science in over 70 countries, both developed and developing ones.

PISA collects data through a two-stage stratified sampling design. With the help of official governments, PISA randomly chooses 150 schools in each country, where they then randomly pick thirty 15 year olds to undertake the two hour tests. The sample size for PISA 2000 is 127,388, for PISA 2003, 276,165, for PISA 2006, 398,750, for PISA 2009, 515,958, for PISA 2012, 480,810, and for PISA 2015 519,334. Together with the subject tests, PISA collects personal information from students, their families and their school environment (including teacher surveys), that serves as relevent background information that can be matched to the students performance. With the recent inclusion of PISA 2015, these six waves make up a time-series analysis of 15 years, enough time to visualize changes in the structure of an educational system.

In order to measure family background PISA includes several variables. Firstly, the ask student's their parent's educational level. Scholars have considered this to be a reliable recall given that we expect fifteen year olds to know their parent's level of education \citep{reardon2011}. This question has been asked in every wave and holds a somewhat similar coding across time, although the first two waves have a different coding. Moreover, another limitation is the fact that parent's education is measured with the ISCED classification, something that has changed over time. For example, until PISA 2009, the ISCED 1997 classification was currently used, but the next wave already used the new ISCED 2011 classification. Both these classification schemes have equivalent look-up tables, yet this requires a detailed inspection of the codings.

Another social background variable PISA collects is the International Socio-Economic Index of Occupational Status (ISEI). This variable captures the economic status of the family, the closest it can without asking for income information. It has been scaled for comparability between waves and some authors have used it for inequality studies finding expected results to be consistent with social origin literature \citep{anna2016_global}. PISA also includes a plethora of indicators on family wealth, home educational resources, the number of books in the home, among many other material resources in the household.

But one of the most relevant variables for our study is the is a composite SES index created by the PISA team. The index of economic, social and cultural status (ESCS) was created on the basis of the following variables: the International Socio-Economic Index of Occupational Status (ISEI); the highest level of education of the student’s parents, converted into years of schooling; the PISA index of family wealth; the PISA index of home educational resources; and the PISA index of possessions related to "classical" culture in the family home \citep{oecd_glance_2002}. This variable, aside from capturing all relevant dimensions of SES, such as education, occupation, and material resources, takes care of transforming the variables into comparable measures across waves. In fact, we use the newly-released ESCS index \citep{pisa_2015_results} which rescaled all ESCS indexes to make them suitable for over-time analysis \footnote{These rescaled indices can be found in the \href{http://www.oecd.org/pisa/data/2015database/}{PISA website} under \emph{Rescaled Indices for Trend Analyses}.}.

Aside from SES, the other most relevant variables are test scores. PISA provides plausible values instead of one single indicator of test score performance. As explained in the PISA manual \citep{pisa2012_technical}, these are imputed values that resemble individual test scores and have approximately the same distribution as the latent trait being measured. Suppose we have \(\mu_i\), the average student test score in mathematics for student \(i\). Instead of estimating \(\mu_i\) alone, plausible values estimate a distribution of possible \(\mu\text{'s}\), for student \(i\), together with the likelihood of each \(\mu_i\) based on the respondents answers on the test. This is defined as the posterior distributions of \(\mu\text{'s}\) for student \(i\). The reason why we use this procedure is because estimating a single estimate \(\mu_i\) is plagued with measurement error, among other types of bias \citep[see][]{wu2005}. The plausible values in the PIAAC data are effectively five (ten for PISA 2015) random draws from this distribution. These variable is are continuous, ranging from 0 to 500, with a mean of 250.

@@

Standardizing the median test score solves the problem of comparability of gaps measured with different tests, and across waves. However, if the variance of academic achievement changes over time, then standardizing the overall score at each wave of testing confounds changes in the gap with changes in the variance of test scores. That is, by standardizing we're forcing the variance of test scores to be zero across all waves. But if the true variance of academic achievement grows over time, then the estimated trend in the achievement gaps will be underestimated, and vice versa.

@@

Inspection of the data suggests that it's something we shouldn't be deeply concerned with. The standard deviation of each wave seems to be following a very similar pattern with not so drastic exception of the year 2000.

>>>>

To investigate the above mentioned questions I will be using the Programme for International Student Assessment (PISA). PISA is a survey carried out every three years that aims to evaluate education systems by testing the skills and knowledge of 15-year-old students. Currrently, PISA has six waves starting which started in 2000 up until 2015, where recently, over half a million students were tested in mathematics, literacy and science in over 70 countries, both developed and developing ones.

PISA collects data through a two-stage stratified sampling design. With the help of official governments, PISA randomly chooses 150 schools in each country, where they then randomly pick thirty 15 year olds to undertake the two hour tests. The sample size for PISA 2000 is 127,388, 276,165 for PISA 2003, 398,750 for PISA 2006, 515,958 for PISA 2009, 480,810 for PISA 2012, and 519,334 for PISA 2015. Together with the subject tests, PISA collects personal information from students, their families and their school environment (including teacher surveys), that serves as relevent background information that can be matched to the students performance. With the recent inclusion of PISA 2015, these six waves make up a time-series analysis of 15 years, enough to visualize changes in the structure of an educational system. None of the literature cited so far has used the last PISA wave, which was recently released in December 2016.

To identify a family's socio-economic status PISA collects several variables that measure different dimensions. Classically, they ask student's their parent's educational level. Scholars have considered this to be a reliable recall given that we expect fifteen year olds to know their parent's level of education \citep{reardon2011}(cite more here). This question has been asked in every wave and holds a somewhat similar coding across time, although the first two waves have a different coding. In spite of this, another limitation is the fact that parent's education is measured with the ISCED classification, something that has changed over time. For example, until PISA 2009, the preferred framework was ISCED 1997, whereas the next wave switched to the newly developed ISCED 2011 classification. Both these classification schemes have equivalent look-up tables, but this requires a detailed inspection of the codings. Aside from all these limitations, this is very reliable and important variable that measures an important factor of socio-economic status.

Another social background variable PISA collects is the International Socio-Economic Index of Occupational Status (ISEI). This variable captures the economic status of the family, the closest it can without asking for income information.  This index variable was developed by Ganzeboom (cite) and it attempts to measure occupational status using a continuous measure. This indicator has been found to ourperform classical occupational classifications such as the Erikson-Goldthorpe-Portocarero schema (cite). It has been scaled for comparability between waves and some authors have used it for inequality studies finding expected results to be consistent with social origin literature \citep{anna2016_global}. PISA also includes a plethora of indicators on family wealth, home educational resources, the number of books in the home, among many other material resources in the household.

Yet one of the most relevant variables for our study is a composite SES index created by the PISA team. The index of economic, social and cultural status (ESCS) was created on the basis of the following variables: the International Socio-Economic Index of Occupational Status (ISEI), the highest level of education of the student’s parents, the PISA index of family wealth (which measured the material wealth of the family), the PISA index of home educational resources; and the PISA index of possessions related to "classical" culture in the family home (mainly about books in the household) \citep{oecd_glance_2002}. This variable, aside from capturing all relevant dimensions of SES, such as education, occupation, and material resources, takes care of transforming the variables into comparable metrics across waves. To the best of our knowledge, this is the first paper that uses the newly-released ESCS index \citep{pisa_2015_results} which was rescaled so that all ESCS indexes are suitable for over-time analysis \footnote{These rescaled indices can be found in the \href{http://www.oecd.org/pisa/data/2015database/}{PISA website} under \emph{Rescaled Indices for Trend Analyses}.} In other words, the ESCS index does not need any transformation or coding updates because it is ready for comparison over time.

Aside from SES, the other most relevant variables are test scores for Mathematics, Literacy and Science. PISA does not provide a single test result for each respondent. Instead, it provides a \emph{series} of 'plausible values' that the child could actually score. As explained in the PISA manual \citep{pisa2012_technical}, these are imputed values that resemble individual test scores and have approximately the same distribution as the latent trait being measured (the true distribution of the possible scores a student can achieve). Suppose we have \(\mu_i\), the average student test score in Mathematics for student \(i\). Instead of estimating \(\mu_i\) alone, plausible values estimate a distribution of possible \(\mu\text{'s}\) for student \(i\), together with the likelihood of each \(\mu_i\) based on the respondents answers on the test. This is defined as the posterior distributions of \(\mu\text{'s}\) for student \(i\). The reason why we use this procedure is because estimating a single estimate \(\mu_i\) is plagued with measurement error, among other types of bias \citep[see][]{wu2005}. The number of plausible values for PISA waves are usually five (although ten for PISA 2015) random draws from this distribution. In practice each student has 5 scores for each test, that resembles this distribution. They are continuous, ranging from 0 to 500, with a mean of 250.

@@

This new standardized variable has a mean of zero. Standardizing the median test score solves the problem of comparability of gaps measured with different tests, and across waves. However, if the variance of academic achievement changes over time, then standardizing the overall score at each country-wave pair actually makes the variable more biased. That is, by standardizing we're forcing the standard deviation of test scores to be zero across all waves. But if the true deviations of the median academic achievement changes over time, then the estimated trend in the SES gaps will be underestimated, or vice versa.

@@

{\centering \includegraphics[width=5.5in,height=5in]{figure/variance_pisa-1}

Inspection of the data suggests that it's something we shouldn't be deeply concerned with. The standard deviation of each wave seems to be following a very similar pattern with a not so drastic exception of the year 2000.
